Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2017-04-25 manuscript.tex
latest

Evolutionary models differ greatly in their treatment of fitness. In models of genetic evolution, genotypes are typically assigned constant (or frequency-dependent) selection coefficients describing the change in their relative frequencies over time. This simplified treatment of selection facilitates genetic realism, and can be justified over sufficiently short time intervals \citep[p. 276]{ewens_2004}. The emphasis here is to predict how allele frequencies change over time based on their fitness effect, genetic drift and linkage, or to infer past selection, drift and linkage given a sample of nucleotide sequences. The resulting picture of evolution excludes basic elements of the ecological underpinnings of selection, including how selection affects population size, and density-dependence. This complicates the inference of past selection, because demographic changes can look geneaologically very similar to selective frequency changes \citep{barton_1998}. Separately fitting demographic models to data on ostensibly neutral sites, and then inferring selection against this demographic background, has had mixed success due to linkage \citep{schrider_2016}.

By contrast, models of phenotypic trait evolution use absolute fitness functions to describe how some traits of interest affect survival and reproduction in particular ecological scenarios. These fitness functions can be quite problem-specific and often only account for a few traits at a time. The emphasis here is on invasion from low frequencies and co-existence, rather than frequency or abundance trajectories over time. For instance, adaptive dynamics uses ``invasion fitness'' to explore the consequences of eco-evolutionary feedbacks \citep{diekmann_2004}.

Less work has been done to generalize beyond particular traits or ecological scenarios to model fundmentally different forms of selection. Perhaps this is not surprising given that fitness is such a complex quantity, dependent on all of a phenotype's functional traits \citep{violle_2007} and its environment. A detailed, trait-based, predictive model of fitness would be enormously complicated and situation-specific. It is therefore easy to doubt the feasibility of a simplified, general mathematical treatment of fitness \citep[p. 276]{ewens_2004}. For example, MacArthur's famous $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}. The r/K scheme's original purpose was to extend the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}, but few attempts have been made to develop it further along these lines.

Empirical trait classification studies have suggested the existence of a few ``primary strategies'', reflecting broadly distinct responses to selection \citep{winemiller_2015}. Grime famously considered \citep{grime_1974,grime_1977,grime_1988,westoby_1998} two broad determinants of population density: stress (persistent hardship e.g. due to resource scarcity or unfavorable temperatures) and disturbance (intermittent destruction of vegetation e.g. due to trampling, herbivory, pathogens, extreme weather or fire).  The extremes of these two factors define three primary strategies denoted by C/S/R respectively (Fig. \ref{fig:grimeschematic}): competitors ``C'' excel in low stress, low disturbance environments; stress tolerators ``S'' excel in high stress, low disturbance environments; and ruderals ``R''  excel in low stress, high disturbance environments. Survival is not possible in high-stress, high-disturbance environments. Grime showed that measures of C, S and R across a wide range of plant species are anti-correlated, so that strong C-strategists are weak S and R strategists, and so on, creating a triangular C/S/R ternary plot \citep{grime_1974}. Similar schemes were proposed for insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, and zooplankton \citep{allan_76}. More recently, modern hierarchical clustering techniques have revealed distinct trait clusters in corals analogous to Grime's primary strategies \citep{darling_2012}. These empirical findings suggest that functional traits contribute to fitness predominantly via a few key factors such as intrinsic reproductive rate or stress-tolerance.

\begin{figure}
\centering
\includegraphics[scale=1]{grimeschematic.pdf}
\caption{\label{fig:grimeschematic} Schematic of Grime's triangle. The two axes show increasing levels of environmental stress and disturbance, respectively. Survival is not possible if the combination of stress and disturbance is too large (dashed line). This creates a triangle, each corner of which corresponds to a ``primary strategy''.} 
\end{figure}

Here we explore the interplay between some ``key factors'' of fitness in a simplified, territorial model of growth, dispersal and competition. This broadly follows the original spirit of MacArthur's $r/K$ scheme, but more specifically, our aim is to begin adding some ecological realism to population genetics' time-depedendent, genetically-focused view of evolution. We revisit the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role, but one critical flaw that we address here.

@@

We assume that reproductively mature individuals (``adults'') each require their own territory to survive and reproduce (Fig.~\ref{fig:lottery}). All territories are identical, and the total number of territories is $T$. Time $t$ advances in discrete iterations, each representing the average time from birth to reproductive maturity. In iteration $t$, the number of adults of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=T-N(t)$. We assume that the $n_i$ and $T$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored. We derive deterministic equations for the expected change in the $n_i$ over time, leaving the evaluation of drift for future work.

@@

In our extension of the classic lottery model, we do not restrict ourselves to high propagule densities. Eq. \eqref{eq:lottery} is nonsensical at low densities ($l_i\ll 1$): genotype $i$ can win at most $m_i$ territories, yet Eq. \eqref{eq:lottery} demands $c_i l_i/\sum_j c_j l_j$ of the $U$ unoccupied territories, for any value of $U$. Intuitively, the cause of this discrepancy is that individuals are discrete. Genotypes with few propagules depend on the outcome of contests in territories where they have at least one propagule present, not some small fraction of a propagule as would be implied by low propagule density $l$ in the classic lottery model. In other words, deviations from the mean propagule composition $l_1,l_2,\ldots,l_G$ are important at low density.

@@

We now discuss the evolutionary change in $b, c$ and $d$ under different environmental conditions. Of specific interest are Grime's ``disturbed'', ``stressful'' and ``ideal'' environments. To proceed, we need to map these verbally-defined environments to quantitative parameter regimes in our model.

The ideal environment is characterized by the near-absence of stress and disturbance: $d_i\ll 1$, whereas $b_i$ is potentially much larger than $1$. From Eq. \eqref{eq:deltN}, the equilibrium value of $L$ only depends on the ratio of birth and death rates. For one genotype, $L/(1-e^{-L})=b_i/d_i$, and so the propagule density is high $L\approx b_i/d_i\gg 1$, and every unoccupied territory will be heavily contested. The population density is also high $N/T\approx 1$ (since $L=b_iN/(N-T)=b_i/(1-T/N)$).

Disturbed environments are characterized by unavoidably high extrinsic mortality caused by physical destruction. Environmental variability, and disturbances in particular, can be modeled as stochastic fluctuations in $b$, $c$ and $d$ \citep{chesson_1981}. For simplicity, we do not pursue this more complicated stochastic approach. Instead, we represent disturbance by high constant mortality rates $d_i$, and extend this mortality to juveniles in the process of territorial contest (since disturbances do not only affect adults as in Eq. \eqref{eq:delttot}). We assume that disturbances are equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each iteration. Disturbed environments then correspond to $d_i$ being close to $1$ for all genotypes (almost all adults and juveniles are killed each iteration). From Eq. \eqref{eq:deltN}, the single genotype equilibrium is given by $L/(1-e^{-L})=d_i/[(1-d_i)b_i]$, and since $L\ll 1$ and $N/T\ll 1$ due to high mortality, we have $L\approx 2(1-d_i/[(1-d_i)b_i])$. Clearly $b_i$ must be exceptionally large to ensure population persistence. The terms proportional to $c_i/\overline{c}$ in Eq. \eqref{eq:master} are then negligible, and $\Delta_+ n_i$ depends primarily on $b_i$.

>>>>

``...the concept of fitness is probably too complex to allow of a useful mathematical development. Since it enters fundamentally into many population genetics considerations, it is remarkable how little attention has been paid to it.'' --- Warren J. Ewens, Mathematical Population Genetics I, 2004

Evolutionary models differ greatly in their treatment of fitness. In models of genetic evolution, genotypes are typically assigned constant (or frequency-dependent) selection coefficients describing the change in their relative frequencies over time. This simplified treatment of selection facilitates genetic realism, and can be justified over sufficiently short time intervals \citep[p. 276]{ewens_2004}. The emphasis here is to infer past selection, migration and demographic change given a sample of nucleotide sequences, or to predict how allele frequencies change over time based on their fitness effects, population structure, genetic drift and linkage. The resulting picture of evolution excludes basic elements of the ecological underpinnings of selection, including density dependence, and how selection affects population size. This complicates the inference of past selection, because demographic changes can look geneaologically very similar to selective frequency changes \citep{barton_1998}.

%Separately fitting demographic models to data on ostensibly neutral sites, and then inferring selection against this demographic background, has had mixed success due to linkage \citep{schrider_2016}.

By contrast, models of phenotypic trait evolution use absolute fitness functions to describe how some traits of interest affect survival and reproduction in particular ecological scenarios \citep{metz_1992,diekmann_2004}. These fitness functions can be quite problem-specific and often only account for a few traits at a time. The emphasis here is on the conditions for invasion from low frequencies and co-existence, rather than frequency or abundance trajectories over time. For instance, adaptive dynamics uses ``invasion fitness'' to explore the consequences of eco-evolutionary feedbacks \citep{diekmann_2004}.

It is challenging to generalize beyond particular traits or ecological scenarios to model fundamentally different forms of selection. Perhaps this is not surprising given that fitness is such a complex quantity, dependent on all of a phenotype's functional traits \citep{violle_2007} and its environment. A detailed, trait-based, predictive model of fitness would be enormously complicated and situation-specific. It is therefore easy to doubt the feasibility of a simplified, general mathematical treatment of fitness \citep[p. 276]{ewens_2004}. For example, MacArthur's famous $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}. However, the r/K scheme's original purpose was to extend the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}. Few attempts have been made to develop it further along these lines.

Empirical trait classification studies have suggested the existence of a few ``primary strategies'', reflecting broadly distinct responses to selection \citep{winemiller_2015}. Grime famously considered \citep{grime_1974,grime_1977,grime_1988,westoby_1998} two broad determinants of population density: stress (persistent hardship e.g. due to resource scarcity or unfavorable temperatures) and disturbance (intermittent destruction of vegetation e.g. due to trampling, herbivory, pathogens, extreme weather or fire).  The extremes of these two factors define three primary strategies denoted by C/S/R respectively (Fig. \ref{fig:grimeschematic}): competitors ``C'' excel in low stress, low disturbance environments; stress tolerators ``S'' excel in high stress, low disturbance environments; and ruderals ``R''  excel in low stress, high disturbance environments. Population persistence is not possible in high-stress, high-disturbance environments. Grime showed that measures of C, S and R across a wide range of plant species are anti-correlated, so that strong C-strategists are weak S and R strategists, and so on, creating a triangular C/S/R ternary plot \citep{grime_1974}. Similar schemes were proposed for insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, and zooplankton \citep{allan_76}. More recently, modern hierarchical clustering techniques have revealed distinct trait clusters in corals analogous to Grime's primary strategies \citep{darling_2012}. These empirical findings suggest that functional traits contribute to fitness predominantly via a few key factors such as intrinsic reproductive rate or stress-tolerance.

\begin{figure}
\centering
\includegraphics[scale=1]{grimeschematic.pdf}
\caption{\label{fig:grimeschematic} Schematic of Grime's triangle. The two axes show increasing levels of environmental stress and disturbance, respectively. Population persistence is not possible if the combination of stress and disturbance is too large (dashed line). This creates a triangle, each corner of which corresponds to a ``primary strategy''.} 
\end{figure}

Here we explore the interplay between some ``key factors'' of fitness in a simplified, territorial model of growth, dispersal and competition. This broadly follows the original spirit of MacArthur's $r/K$ scheme. More specifically, our aim is to begin adding some ecological realism to population genetics' time-depedendent, genetically-focused view of evolution. We revisit the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role, but one critical flaw that we rectify here.

@@

We assume that reproductively mature individuals (``adults'') each require their own territory to survive and reproduce (Fig.~\ref{fig:lottery}). All territories are identical, and the total number of territories is $T$. Time $t$ advances in discrete iterations, each representing the time from birth to reproductive maturity. In iteration $t$, the number of adults of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=T-N(t)$. We assume that the $n_i$ and $T$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored. We derive deterministic equations for the expected change in the $n_i$ over time, leaving the evaluation of drift for future work.

@@

In our extension of the classic lottery model, we do not restrict ourselves to high propagule densities. Eq. \eqref{eq:lottery} is nonsensical at low densities ($l_i\ll 1$): genotype $i$ can win at most $m_i$ territories, yet Eq. \eqref{eq:lottery} demands $c_i l_i/\sum_j c_j l_j$ of the $U$ unoccupied territories, for any value of $U$. Intuitively, the cause of this discrepancy is that individuals are discrete. Genotypes with few propagules depend on the outcome of contests in territories where they have at least one propagule present, not some small fraction of a propagule as would be implied by small $l_i$ in the classic lottery model. In other words, deviations from the mean propagule composition $l_1,l_2,\ldots,l_G$ are important at low density.

@@

Here we describe how Grime's ``disturbed'', ``stressful'' and ``ideal'' environments can be captured mathematically in our scheme. To proceed, we need to map these verbally-defined environments to quantitative parameter regimes in our model.

The ideal environment is characterized by the near-absence of stress and disturbance: $d_i\ll 1$, whereas $b_i$ is potentially much larger than $1$.

%From Eq. \eqref{eq:deltN}, the equilibrium value of $L$ only depends on the ratio of birth and death rates. For one genotype, $L/(1-e^{-L})=b_i/d_i$, and so the propagule density is high $L\approx b_i/d_i\gg 1$, and every unoccupied territory will be heavily contested. The population density is also high $N/T\approx 1$ (since $L=b_iN/(N-T)=b_i/(1-T/N)$).

Disturbed environments are characterized by short bursts of high elevated extrinsic mortality caused by physical destruction

high extrinsic mortality , or, more accurately, .

In the disturbance extreme of Grime's triangle we can

, we instead represent disturbance by high constant mortality rates $d_i$, and extend this mortality to juveniles in the process of territorial contest (since disturbances do not only affect adults as in Eq. \eqref{eq:delttot}).

We assume that disturbances are equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each iteration. Disturbed environments then correspond to $d_i$ being close to $1$ for all genotypes (almost all adults and juveniles are killed each iteration).

%From Eq. \eqref{eq:deltN}, the single genotype equilibrium is given by $L/(1-e^{-L})=d_i/[(1-d_i)b_i]$, and since $L\ll 1$ and $N/T\ll 1$ due to high mortality, we have $L\approx 2(1-d_i/[(1-d_i)b_i])$. Clearly $b_i$ must be exceptionally large to ensure population persistence. The terms proportional to $c_i/\overline{c}$ in Eq. \eqref{eq:master} are then negligible, and $\Delta_+ n_i$ depends primarily on $b_i$.
